Human capital mediates natural selection in contemporary humans

### 2021-08-09 why-natural-selection.Rmd
Final tweaks pre-NHB submission

2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
    at higher wages and/or levels of                  polygenic scores for EA (Appendix Figure
    human capital.                                    \@ref(fig:plot-purifying)).

@@

One interpretation of these results is that natural selection on polygenic
scores is driven by their correlation with human capital, which affects
fertility as our model describes. This does not necessarily imply that polygenic
scores are causal of either human capital or fertility. Instead they may simply
correlate with different family environments, for example via "genetic nurture"
effects [@kong2018nature] or population stratification. To test whether
polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
among full siblings, controlling for a full set of sibling group dummies
(Appendix Figure \@ref(fig:plot-res-fe-fertility)). All within-sibling effects
are insignificant after Bonferroni correction. Effect sizes are positively
correlated with effect sizes from the pooled model, and about 1/3 smaller
(regressing within-sibling on pooled effect sizes, b = `r coef_pool_w`, se = 
`r se_pool_w` ). Thus, there is only weak evidence that polygenic scores are
causally linked to fertility, and selection effects are at least partially
driven by gene-environment correlation. Also, if we add a control for education 
to the within-sibling regressions, effect sizes barely change. This
suggests that our measure of education only mediates differences in fertility
between sibling groups; among siblings, other causal channels are more
important.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$ in `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In addition, genetic effects on
educational attainment have been shown to be inflated in population-based
samples as compared to within-family designs, likely because of indirect genetic 
effects, gene-environment correlations, and/or assortative mating 
[@lee2018gene; @selzam2019comparing; @kong2018nature]. 
In short, it is probably too early to tell whether modern natural selection has a
substantively important effect on the genetic make-up of the population.
Nevertheless, we note that selection effects on our measured polygenic scores
are still relatively small, even after reweighting to account for ascertainment
bias.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018], then these increases are substantial effects. Since many
polygenic scores are predictive of disease risk, they could also potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

>>>>

2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
    at higher wages and/or levels of                  polygenic scores for 
    human capital.                                    educational attainment (Appendix Figure
                                                      \@ref(fig:plot-purifying)).

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. This need not imply
that a given polygenic score is causal of either fertility or human capital. Instead
scores may simply correlate with causally relevant environments, for example via
"genetic nurture" effects [@kong2018nature] or population stratification. To
test whether polygenic scores cause fertility, we estimate equation
\@ref(eq:regression) among full siblings, controlling for a full set of sibling
group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
sample size, all within-sibling effects are insignificant after Bonferroni
correction. However, effect sizes are positively correlated with effect sizes
from the pooled model, and about 1/3 smaller (regressing within-sibling on
pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
with the ~50% decrease in heritability in within-sibling GWASs on age at first
birth and educational attainment [@Howe2021.03.05.433935]. We see these results
as providing weak evidence that polygenic scores cause fertility, with effects
being at least partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores (EA2 and EA3), effect sizes are reduced by about 20%.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and younger age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$ in `r n_risk_sig` out of 33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In addition, genetic effects on
educational attainment have been shown to be inflated in population-based
samples as compared to within-family designs, likely because of indirect genetic 
effects, gene-environment correlations, and/or assortative mating 
[@lee2018gene; @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935]. 
In short, it is probably too early to tell whether modern natural selection has a
substantively important effect on the genetic make-up of the population.
Nevertheless, we note that selection effects on our measured polygenic scores
are still relatively small, even after reweighting to account for ascertainment
bias.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
important for understanding class mobility and structure [@belsky2018genetic;
@Rimfeld_2018], then these increases are substantially important. Since many
polygenic scores are predictive of disease risk, they could also potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.
